---
title: "Recommendation Systems - Collaborative Filtering(User Similarity)"
author: "Tarun Reddy Aleti"
date: "April 8, 2016"
output: html_document
---
### Executive Summary
- The datasets required for the analysis can be downloaded from these links [Movies](https://inclass.kaggle.com/c/movie/download/movies.dat), [Users](https://inclass.kaggle.com/c/movie/download/users.dat), [Ratings](https://inclass.kaggle.com/c/movie/download/training_ratings_for_kaggle_comp.csv) and [ReadMe](https://inclass.kaggle.com/c/movie/download/README) 
- Idea is to use Weighted K-Nearest Neighbours Machine Learning Algorithm to develop a recommendation system for movies for a given user

### Required Packages

Following packages are required for the code to run efficiently.

```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
library(MASS)
library(plyr)
library(knitr)
library(caret)
library(sqldf)
```

### Preliminary Analysis

```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
set.seed(1)
options(scipen = 999)
#In MAC use encoding latin1
movieLines <- readLines("movies.dat")
movies <- adply(movieLines, 1, function(x){
  y <- strsplit(x, split = ":")
  return(c(y[[1]][1], y[[1]][3], y[[1]][5]))
})
movies <- movies[ ,-1]
colnames(movies) <- c("MovieID", "Title", "Genres")
users <- read.delim("users.dat", header = FALSE, sep = ":", stringsAsFactors = FALSE)
users <- users[ ,seq(1, 9, 2)]
colnames(users) <- c("UserID", "Gender", "Age", "Occupation", "Zipcode")
ratings <- read.csv("training_ratings_for_kaggle_comp.csv", header = TRUE, stringsAsFactors = FALSE)

genres <- c("Action", "Adventure", "Animation", "Children's",
            "Comedy", "Crime", "Documentary", "Drama", "Fantasy",
            "Film-Noir", "Horror", "Musical", "Mystery", "Romance",
            "Sci-Fi", "Thriller", "War", "Western")


ages <- levels(as.factor(users$Age))
occupations <- levels(as.factor(users$Occupation))
genders <- levels(as.factor(users$Gender))
```

- We have two datasets in hand - Movies and Ratings
  1. *Movies*: It has three fields MovieID, MovieTitle and Genres.
    + We can extract the year movie was released using the title name. We can further extract the decade from the year.
    + Movie can belong to more than one of the Genre. We need to create a dummy variable indicating if the movie belong to that Genre or not.
  2. *Ratings*: It has four field UserID, MovieID, Rating and User_Movie_ID
    + All the fields are self explanatory.
    + We need to join all these tables and Rating is our dependent variable for our model.
  3. *Users*: It has five fields UserID, Gender, Age, Occupation and Zipcode
    + We can use Zipcode information and extract the city, state the Users belong to.
    + Age is already categorized into 6 categories and Occupation into 20 categories as can be from Readme file.
    
### Feature Extraction
- Create dummy variables for each genre
- Combine the 3 data sets to come with a single dataset
- We could create dummy variables for year movie released (decade it belongs to) but the Movie Name column is not clean. Even after multiple trials we couldn't find a way out. We are ignoring this feature for now.
- We create dummy variables for users age, gender and occupation.
- Just like movie titles even after lot of cleaning zip codes were incorrect, hence, for now we will ignore them.

```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
movie_genre <- adply(movies[ ,'Genres'], 1, .id = NULL, function(x){
  dummy_genre <- rep(0, length(genres))
  g <- strsplit(x, split = '[|]')
  index_g <- aaply(g[[1]], 1, function(x){
    return(which(genres == x))
  })
  dummy_genre[index_g] <- 1
  return(dummy_genre)
})
colnames(movie_genre) <- paste("genres", genres, sep = "_")

users_age <- adply(users[ ,'Age'], 1, .id = NULL, function(x){
  dummy_age <- rep(0, length(ages))
  dummy_age[which(ages == x)] <- 1
  return(dummy_age)
})
colnames(users_age) <- paste("ages", ages, sep = "_")

users_occupation <- adply(users[ ,'Occupation'], 1, .id = NULL, function(x){
  dummy_occupation <- rep(0, length(occupations))
  dummy_occupation[which(occupations == x)] <- 1
  return(dummy_occupation)
})
colnames(users_occupation) <- paste("occupations", occupations, sep = "_")

users_gender <- adply(users[ ,'Gender'], 1, .id = NULL, function(x){
  dummy_gender <- rep(0, length(genders))
  dummy_gender[which(genders == x)] <- 1
  return(dummy_gender)
})
colnames(users_gender) <- paste("genders", genders, sep = "_")

users <- do.call("cbind", list(users, users_gender, users_age, users_occupation))
users <- users[ ,-c(2:5)]
movies <- cbind(movies, movie_genre)
rownames(movie_genre) <- movies$MovieID

query <- "SELECT m.*, u.*, r.rating
          FROM ratings as r
          INNER JOIN movies as m on r.movie = m.MovieID
          INNER JOIN users as u on r.user = u.UserID"

data <- sqldf(query)
data$UserID <- as.character(data$UserID)
```

### Train and Test Sets
- We need to create a train and test from the data set
- We shall assign 10% of data randomly to test and 90% to train
- We can finally test the algorithm on kaggle test data set.

```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
slctd_users <- sample(unique(data$UserID), length)
data <- data[ ,c(setdiff(colnames(data), c('Genres')))]
data <- data[!is.na(match(data$UserID, slctd_users)), ]
trainIndex <- createDataPartition(data$rating, p = .7, list = FALSE)
train <- data[trainIndex, ]
test <- data[-trainIndex, ]




trainIndex <- createDataPartition(data$UserID, p = .9, list = FALSE)
train <- data[trainIndex, ]
test <- data[-trainIndex, ]
finaltest <- read.csv("sample_submission.csv", header = TRUE)
finaltest$movie <- apply(finaltest, 1, function(x){
  return(strsplit(as.character(x[3]), split = "_")[[1]][2])
})  


query <- "SELECT m.*, u.*, f.rating
          FROM finaltest as f
          INNER JOIN movies as m on f.movie = m.MovieID
          INNER JOIN users as u on f.user = u.UserID"

finaltest <- sqldf(query)

finaltest[is.na(finaltest$Gender), 'Gender'] <- 'NA'
finaltest$Age <- as.character(finaltest$Age)
finaltest$Occupation <- as.character(finaltest$Occupation)
finaltest[is.na(finaltest$Age), 'Age'] <- 'NA'
finaltest[is.na(finaltest$Occupation), 'Occupation'] <- 'NA'
finaltest$UserID <- as.character(finaltest$UserID)
```

### Utility Matrix

- Utility matrix consists for all the ratings given by a user for a given movie.
- If the user hasn't seen the movie then rating would be 0
- We use train data set to create the utility matrix

```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
set.seed(1)
shrt_users <- sort(users$UserID)
shrt_movies <- sort(movies$MovieID)
ut_mat <- data.frame(matrix(0, ncol = length(shrt_movies), nrow = length(shrt_users)))
rownames(ut_mat) <- shrt_users
colnames(ut_mat) <- shrt_movies

#Fill the utility matrix with ratings given by userid for a movieid
ut_mat <- adply(ut_mat, 1, .id = NULL, function(x){
  y <- train[train$user == rownames(x), ]
  z <- y[match(colnames(x), y[ ,'MovieID']), 'rating']
  z[is.na(z)] <- 0
  return(z)
})
colnames(ut_mat) <- shrt_movies
rownames(ut_mat) <- shrt_users

genres <- gsub("['-]", "", genres)
colnames(movie_genre) <- gsub("['-]", "", colnames(movie_genre))
colnames(train) <- gsub("['-]", "", colnames(train))
colnames(test) <- gsub("['-]", "", colnames(test))
colnames(finaltest) <- gsub("['-]", "", colnames(finaltest))
```

### Item Profile
- First we need to find the average rating given by a user and subtract from the ratings 
- We need to calculate the similarity between users first
	+ We shall use ratings given by user for a movie(m) as well as features that define a user(f)
- To estimate rating a user would have given for a movie we need to find each movie rating given by nearest 10 users to the user who would have
given a rating and then take average of those ratings
	+ **10** is chosen arbitarily can be changed to find the best value.

```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
utility_mat <- adply(ut_mat, 1, .id = NULL, function(x){
  if (length(which(x > 0)) > 0){
    x[which(x > 0)] <- x[which(x > 0)] - mean(as.numeric(x[which(x > 0)]))
  }
  return(x)
})

### Option 1: Finding Similar users to one user. We need to find users that have
#more cosine distance between movies they like
#utility_matrix is u x (m + f) where u is number of users and m is number of movies and f is users features
#We need a matrix u x u which has cosine distances between users

utility_mat <- adply(ut_mat, 1, .id = NULL, function(x){
  if (length(which(x > 0)) > 0){
    x[which(x > 0)] <- x[which(x > 0)] - mean(as.numeric(x[which(x > 0)]))
  }
  return(x)
})

utility_mat <- cbind(utility_mat, users[shrt_users, -1])

utility_mat_mag <- adply(utility_mat, 1, function(x){
  return(x/sqrt(sum(x^2)))
})
utility_mat_mag[is.na(utility_mat_mag)] <- 0

user_user <- as.matrix(utility_mat_mag) %*% as.matrix(t(utility_mat_mag))
colnames(user_user) <- shrt_users
rownames(user_user) <- shrt_users


#For each movie raitng given by a user lets choose nearest 10 users to the user who would have
#given a rating and then take average of those ratings

user_movie_rating <- adply(ut_mat, 1, .id = NULL, function(x){
  #Every column in this a movie
  users_dist <- data.frame(user_user[setdiff(rownames(user_user), rownames(x)), rownames(x)])
  colnames(users_dist) <- 'Distance'
  users_dist$UserID <- setdiff(rownames(user_user), rownames(x))
  #Average rating given by the User
  if(length(which(x > 0)) > 0){
    avg_rating_by_user <- mean(as.numeric(x[which(x > 0)]))
  } else {
    avg_rating_by_user <- 0
  }
  all_movies <- adply(x, 2, function(y){
    #For a given user and movie find the nearest 10 users in user_user who have given a rating
    #for that movie in ut_mat
      all_user_movie_rating <- data.frame(ut_mat[setdiff(rownames(user_user), rownames(x)), colnames(y)])
      colnames(all_user_movie_rating) <- 'rating'
      all_user_movie_rating$UserID <- setdiff(rownames(user_user), rownames(x))
      query <- "SELECT u.UserID, u.Distance, a.rating
                FROM users_dist as u 
                INNER JOIN all_user_movie_rating as a
                ON u.UserID = a.UserID
                WHERE a.rating != 0
                ORDER BY u.Distance desc"
      nearest_user_movie_ratings <- sqldf(query)
      if(dim(nearest_user_movie_ratings)[1] == 0){
        avg_rating_by_near_users <- 0
      } else if (dim(nearest_user_movie_ratings)[1] <= 10){
        avg_rating_by_near_users <- as.numeric(mean(nearest_user_movie_ratings$rating))
      } else {
        nearest_10_user_movie_ratings <- nearest_user_movie_ratings[1:10, ]
        #Average of the 10 ratings in the value of rating for a given user and movie combination
        avg_rating_by_near_users <- as.numeric(mean(nearest_10_user_movie_ratings$rating))
      }
      return(avg_rating_by_near_users)
  })
  all_movies <- data.frame(t(all_movies[ ,2]))
  colnames(all_movies) <- shrt_movies
  return(all_movies)
})
rownames(user_movie_rating) <- shrt_users
```

### Prediction
- We have a matrix which will gives us user rating for a movie
- In order to predict the rating the user would have given to a new movie we need to take following steps
	+ Identify row the user belongs to in matrix
	+ Identify column the user belongs to in matrix
	
```{r, echo = TRUE, warning = FALSE, message = FALSE, results = 'hide'}
test$rating_pred <- adply(test, 1, .id = NULL, function(x){
  row_id <- which(rownames(user_movie_rating) == x$UserID)
  col_id <- which(colnames(user_movie_rating) == x$MovieID)
  if(length(row_id) == 1 & length(col_id) == 1){
    return(user_movie_rating[row_id, col_id])
  } else {
    return(NA)
  }
})
```

### Conclusion
- Successfully used collaborative filtering user similarity method to make a rating prediction for a given user and movie combinations